Efficient Compilation of High Level Python Numerical Programs with Pythran

### 2014-08-28 doc/papers/cgo2014/paper.tex
Add some perfs mesures
More info on forward substitution (not finished yet)

\subsection{Expression Templates}

@@

\begin{lstlisting}[language=c++, caption={C++11 translated version of Python version for the rosenbrock kernel.}, label={lst:rosen-cxx}, breaklines=true]
constexpr size_t N = x::ndim;
ndarray<N> t0 = 100L * numpy::square{}(x[slice(1L, None)] - numpy::square{}(x[slice(None,-1L)]));
ndarray<N> t1 = numpy::square{}(1L - x[slice(None,-1L)]);
return numpy::sum{}((t0 + t1));
\end{lstlisting}

@@

>>>>

\subsection{Expression Templates}
\label{sec:expression-templates}

@@

This version does not supress all intermediate computations: typically, the
assignments to \texttt{t0} and \texttt{t1} trigger the computation of the right
hand side expression, thus creating two intermediates array. This usual
limitation is motivated by the inability to prove that \texttt{x} is not
modified before \texttt{t0} is read.

\begin{lstlisting}[language=c++, caption={C++11 translated version of Python version for the rosenbrock kernel.}, label={lst:rosen-cxx}, breaklines=true]
typedef typename x::dtype T;
constexpr size_t N = x::ndim;
ndarray<T, N> t0 = 100L * numpy::square{}(x[slice(1L, None)] - numpy::square{}(x[slice(None,-1L)]));
ndarray<T, N> t1 = numpy::square{}(1L - x[slice(None,-1L)]);
return numpy::sum{}((t0 + t1));
\end{lstlisting}

@@

The limitation of expression templates presented in
\S~\ref{sec:expression-templates} is easy to solve at the Python compiler
level, using forward substitution. Typically, forward substitution is
supported by the computation of a dependency graph that itself relies on
read-write effects and alias analysis.

Alias analysis is impossible for Python programs in general, because of late
binding and possible monkey patching. In the context of a static compiler for
numerical programs, it can be assumed that binding is done statically and no
monkey patching is done.  In that context, it is relatively easy to design an
intra-procedural alias analysis thanks to the absence of pointer arithmetic:
the only language construction that can create aliasing is the assignment, thus
a simple control-flow base alias analysis yields accurate results.

The computation of read-write effects involve the annotation of numpy functions
so that the Python compiler knows that most of them do not modify their
argument, and the computation of argument effects for user functions. This can
be done by considering each function individually and looking for the two kind
of constructs that can modify a function argument: indexing and further
function calls.

To interprocedurally compute the read-write effects, intraprocedrual effects
are first computed. Then we create a directed graph $G$ with as many nodes as
functions. Each node contains a list of integer representing the indices of the
formal argument that are modified intra-procedurally. Then we add an edge
between two nodes, labeeld by an integer tuple $(i, j)$ if $i^\text{th}$ formal
argument of the source is passed as $j^\text{th}$ effective argument of the
sink. The goal is to prune all edges from $G$. To do so, given a node $n$ that
writes $i$, an outcoming edge $(i,j)$ can be safely removed, as we know for
sure that the $i^\text{th}$ parameter is written.  Similarly, an incoming edge
$(j,i)$ can be removed and $j$ is added to the source's list. Iteratively
applying this process removes all edges that can propagate a write, the only
remaining edges propagate a read and can be ultmately removed.
